# Importing & Summarizing Data in R 

## Goal of Software Lesson 1

::: {.graybox}
The goal of this software lesson is to learn how to

  * import data from a CSV file,  
  
  * examine the contents of the data, and   
  
  * create summary statistics, tables, and graphs     for a variable. 

:::
    




## Dataset

The data set that will be used for the software lessons comes from the Lung Health Study (LHS). 

The LHS was a multicenter randomized clinical trial investigating whether smoking intervention and use of an inhaled bronchodilator (Atrovent) would decrease the rate of decline in lung function over the 5-year follow-up period. A total of 5,887 participants (smokers, aged 35-67 years old) were recruited from 10 clinical centers in the US and Canada between the years of 1986 to 1988. They were randomized into three groups: 

* usual care (UC): advised to stop smoking and followed;  

* special intervention with Atrovent (SIA): enrolled in an intensive smoking cessation program & was prescribed an inhaled bronchodilator (Atrovent); and,  

* special intervention with Placebo (SIP): enrolled in an intensive smoking cessation program & was prescribed a placebo.

Baseline measurements and measurements at each of 5 annual visits were collected on each of the participants.

The data set and a description of the 80 variables can be found on Canvas, titled *lhs.csv* and *lhsdoc.docx*, respectively.

## Importing data

Two ways of importing the data will be presented. The first option uses the `read.csv()` function and can be carried out in either R or R Studio. The second way uses a built-in feature of R Studio, so it can only be carried out in R Studio and not R.

**Option #1**

- To import a CSV data set from your computer into R, use the `read.csv()` function:
\vspace{0.15cm}
```{r, eval=FALSE, echo=TRUE, comment = ""}
lhs<-read.csv(file = file.choose(), header = TRUE)
```

Let's break apart the important pieces in the code above:

* <b> object </b> (e.g., `lhs`), 

* <b> assignment operator </b> (`<-`), and

* <b> function </b> (e.g., `read.csv()`).

The first important part in the above code is saving the data set as a name (or **object**) in R. R uses object-oriented programming, which means that data are assigned to an object and you, the programmer, are in charge of naming that object. You create objects in R to store information for later use. The name or object created in the example above is called `lhs`. A good programmer will name the objects in a meaningful way in order for easy recall of what is stored in that object.  

The next important part is using the **assignment operator** (`<-`) to assign information to the object.

The other important part in the above code is using the appropriate **function** to import the data into R. A function in R allows you to specify some inputs in order to obtain particular outputs. The function that is used to import a CSV file into R is the `read.csv()` function. The inputs that are specified by you are called **arguments** and are separated by commas in the function's parentheses. There are two necessary arguments needed in the `read.csv()` function: `file = ` and `header = `. 
  
The first argument, `file = `, tells R where the CSV file is located, on your computer or on the web. If the file is on your computer, the easiest way to find the file is through the function `file.choose()`. This function allows you to browse your computer for the location of the CSV file. Once found, select the CSV file (*lhs.csv*, in this situation) and click *Open*.  
  
The other argument, `header = `, tells R whether or not the first row of the CSV file contains the variable names. If it does, then you would type `header = TRUE` (as seen in the code above). If the first row does not contain the variable names and the file starts with data values in the first row, then the argument would change to `header = FALSE`.


::: {.redbox}
**R TIP**

R is case sensitive, which means that the symbols <bdi style = "font-family: 'Courier New'"> A </bdi> and <bdi style = "font-family: 'Courier New'"> a </bdi> would not refer to the same thing. Make sure you pay attention to the details of the naming of the functions, arguments of the functions, or objects.

:::
<br>


**Option #2** (R Studio only):   
  
- Go to *File -> Import Dataset -> From Text (base)...*  
  
- Browse your computer for the location of the CSV file (*lhs.csv*, in this situation), select the file, and click *Open*. 

- In the new window, give your data set a name (e.g., lhs), make sure the option for Header is selected **Yes**, and click *Import*.


::: {.bluebox}
**IMPORTING DATA**

<b>Purpose:</b> To read in an external data set into R.
<br>
<div style="border-bottom: 1px dashed black"></div>
<br>
For additional R resources on importing data into R, see: 

* <a href ="https://www.statmethods.net/input/importingdata.html">Importing Data [Quick-R]. </a>
* <a href ="http://www.sthda.com/english/wiki/importing-data-into-r">Importing Data into R [STHDA]. </a>
* <a href ="https://support.rstudio.com/hc/en-us/articles/218611977-Importing-Data-with-RStudio">Importing data with RStudio [RStudio Support]. </a>

:::


## Examine the data

### View the dataset

Once the data has been imported into R, it's good coding practice to view your data set to make sure everything was read in properly.

One way to view your data set is by viewing the first few rows of the data set. 

- To view the first few rows of the data, use the `head()` function, specifying the object or name of the data set:

```{r, eval=FALSE,echo=TRUE, comment = ""}
head(lhs)

### To modify the number of rows presented, use head(lhs, n=10),
### where the argument n= denotes the number of rows to be displayed
```

The function `head()` displays the first six rows of the data set. So if you were to read the line of code above, it would read as "I would like the first 6 rows (i.e., `head()`) of the `lhs` data set". The "of" is denoted by the parentheses around `lhs`.

:::{.redbox}
**R TIP**

You may want to add comments within your code so that you can recall at a later time what was executed. To do this, type on a new line # {this is my comment} and R will ignore everything after the #.

:::

<br>

If you wanted to look at the entire data set at once in the console, you could do so by typing the name of the object (e.g., `lhs`). However, this is unwise for large data sets, like the LHS one. Rather, it is better to look at the data in small chunks as presented above.

There is one exception to the note above. If you are using R Studio, you can view the data in the source window as a spreadsheet. 

- In R Studio, to view the data in the source window, use the `View()` function, specifying the object or name of the data set:

```{r, eval=FALSE, echo=TRUE, comment = ""}
View(lhs)
```


:::{.bluebox}
**VIEWING THE DATASET**

For additional R resources on viewing the data set in R, see: 

* <a href ="http://rfunction.com/archives/699">head, tail [R Function of the Day]. </a>
* <a href="http://www.datasciencemadesimple.com/get-first-n-rows-last-n-rows-head-and-tail-function-in-r/">Get first n rows and last n rows with head and tail function in R [Data Science Made Simple]. </a> <br>
Note: Scroll about half way down the page to see the article on the head and tail function.

:::

### View data types of the variables

Another good coding practice is to examine the data types of the variables in the data set to make sure they are correctly denoted in R.

R has 6 data types, but only 4 will be used in this class. A variable can classified in R as a

* character (or categorical variable), labeled as `chr`: "Male", "High Cholesterol";

* numeric (or continuous variable, if contains decimals), labeled as `num`: 15.5, 20;

* integer (or whole number or discrete variable), labeled as `int`: 0, 1, 2, 3; and

* logical, labeled as `logi`: TRUE, FALSE

<br>

To view the data types for all of the variables in the data set, use the `str()` function, specifying the object or name of the data set:  
```{r, eval=FALSE, echo=TRUE, comment = ""}
str(lhs)
```

::: {.redbox}
**R Tip**  

Variables that are categorical but are recorded as 0/1 (or 1/2) in the data set will be treated as an integer data type (i.e., `int`), even though they are not actually a numerical variable. In some instances, you may have to modify the data type of a variable in R to carry out particular analyses.
:::


<br>


::: {.bluebox}
**VIEWING DATA TYPES**

For additional R resources on viewing data types of the variables in R, see: 

* <a href ="https://swcarpentry.github.io/r-novice-inflammation/13-supp-data-structures">Data Types and Structures [Programming with R]. </a>
* <a href = "https://www.r-bloggers.com/basic-data-types-in-r">Basic Data Types in R [R-Bloggers]. </a> 

:::



### View names of the variables

The last useful tip when examining the data is to view the names of the variables in the data set. This is useful because you want to make sure you are using the correct variable name as R sees it.

- To view the names of all of the variable in the data set, use the `names()` function, specifying the object or name of the data set:  
```{r, eval=FALSE, echo=TRUE, comment = ""}
names(lhs)
```


## GUIDED QUESTIONS {#g11}

##### Question 1

What is the age of the participant in the third row of the data set?

##### Question 2

What data type did R put for the race variable, `race`? For the clinical center code variable, `ACLINIC`? For the BMI variable, `bmi`? 

##### Question 3
Using the LHS data documentation and the output from the `str()` function, do the data types of the variables in R match the actual type for the variables? If they do not match for a variable, think about why R incorrectly defined a data type for the variable. That is, what is it about the way the data was recorded that made R misclassify the data type?

::: {.graybox}

See \@ref(answers-1) for answers

:::



## Summarizing the data

Now that you have the basics of your software (e.g., import a CSV file, examine the data), the next step is to get a sense of the variables' values in the data set by summarizing them (e.g., counts/number and/or percents; measures of center and spread; graphical displays).

### Summarize all variables at once

The fastest way to get a sense of the variables is to summarize them all at once.

- To summarize all of the variables in the data set at once, use the `summary()` function, specifying the object or name of the data set:

```{r, eval=FALSE, echo=TRUE, comment = ""}
summary(lhs)
```

The `summary()` function provides different summaries, depending on the data type of the variable.  

For continuous or integer variables, six summary measures/statistics are computed: minimum (`Min.`), Q1 (`1st Qu.`), median (`Median`), mean (`Mean`), Q3 (`3rd Qu.`), and maximum (`Max.`). 

Unfortunetaly, for categorical variables (that are `chr` types), R doesn't tell you any summary information at all about the variable, apart from the length (or number of rows) and the class type (which is `character`). If the categorical variable is a `Factor` (another data type in R for categorical variables), then the summary measure/statistic is the count for each category in the variable.

The `summary()` function also provides a count of missing values for each variable, as indicated by `NA's   :`. This is only displayed for a variable if there are missing values for that variable.

<br>

::: {.redbox}
**R TIP**

Remember to examine the description of the data (aka data dictionary) to make sure you are interpreting the summaries correctly. Recall that some of the variables may really be categorical but were recorded as integers (e.g., 0's and 1's; 1's and 2's; 1's, 2's, 3's, 4's, and 5's). As a result, R will, by default, summarize the variable as if it was a continuous or integer variable, which would not be correct for the variable. 

:::


## GUIDED QUESTIONS {#g12}

##### Question 4
What is the five-number summary of number of cigarettes per day at baseline (see `f10cigs` variable)? What is the mean number of cigarettes per day at baseline? 

##### Question 5
Which of the following variables have missing values: number of years of education at baseline (`yeareduc`), weight change (kg) from baseline to 2nd annual visit (`wgtchg02`), salivary cotinine at 4th annual visit (`COT4`)?

##### Question 6
Does it make sense to interpret the summaries computed for the `race` variable? Why or why not?

::: {.graybox}

See \@ref(answers-1) for answers

:::

### Summarize single categorical variable

It is good practice to examine the data when you first import it into your software, and summarizing all the variables at once is part of that.  

However, it's most often of interest to focus on a single variable or subsets of the variables. Let's now shift to learning how to summarize single variables in the data set, starting with summarizing a single categorical variable. The LHS variable that will be the focus in this section is the marital status variable, `marital0`.

To summarize categorical variables, one is often interested in the counts/frequencies and proportions within the categories. This can be displayed in a table or in a graph (i.e., bar plot). 

#### Table of counts

- To create a table of counts for a categorical variable, use the `table()` function, specifying the object or name of the data set, followed by a `$`, and then the variable name of interest (e.g., `lhs$marital0`):

```{r, eval=FALSE, echo=TRUE, comment = ""}
table(lhs$marital0)
```

This function produces a count for each unique category in the variable. 
<br>

::: {.redbox}
**R TIP**

Whenever you want to look at a particular variable in the data set (e.g., <bdi style = "font-family: 'Courier New'"> marital0</bdi>), the structure of the R code will be <bdi style = "font-family: 'Courier New'"> object\$variable\_name </bdi> (e.g.,  <bdi style = "font-family: 'Courier New'"> lhs\$marital0</bdi>). 

:::

#### Table of proportions

If you are interested in creating a table of proportions instead of counts, you first need to save the count information from the `table()` function into an object (e.g., `counts`) and then use that object in the function for calculating the proportions for each unique category (`prop.table()`).

- Step 1: To save the count information from the `table()` function into an object, first, name the object something meaningful (e.g., `counts`), and then use the assignment operator (`<-`) to assign the table of counts (e.g., `table(lhs$marital0)`) to the object:

```{r, eval=FALSE, echo=TRUE, comment = ""}
###Save the count information from the table() function
counts<-table(lhs$marital0)
```

- Step 2: To create a table of proportions for a categorical variable, use the `prop.table()` function and specify the saved count object (e.g., `counts`): 

```{r, eval=FALSE, echo=TRUE, comment = ""}
###Create a table of proportions using the counts object
prop.table(counts)
```

This function produces a proportion for each unique category in the variable. 

#### Bar plot

When creating a bar plot for a categorical variable, the instructions are similar to that for creating a table of proportions: you first need to save the count information from the `table()` function into an object (e.g., `counts`) and then use that object in the function for bar plots (`barplot()`). Because you created the `counts` object previously, that step will not be shown again and will go straight into using the object in the bar plot function.

- To create a bar plot, first, use the `barplot()` function, specifying the saved count object (e.g., `counts`) and other arguments if you want to customize the bar plot (e.g., `main=` for the plot title and `names.arg=` for the bar labels):

```{r, eval=FALSE, echo=TRUE, comment = ""}
###Create a barplot using the counts object
barplot(counts, main="Barplot of LHS Marital Status", 
        names.arg=c("Never","M","W","S","D"))
```

::: {.redbox}
**R TIP**

If you want the formal R documentation for a function, type <bdi style = "font-family: 'Courier New'"> ?</bdi> before the function name (e.g., <bdi style = "font-family: 'Courier New'"> ?barplot</bdi>). This is useful when you want to see what other arguments can be called in the function.

:::


<br>


::: {.bluebox}
**SUMMARIZE SINGLE CATEGORICAL VARIABLE**

<b>Purpose:</b> To summarize a single categorical variable using summary measures and graphical representations.
<br>
<div style="border-bottom: 1px dashed black"></div>
<br>
For additional R resources on summarizing categorical variables, see: 

* <a href="http://www.sthda.com/english/wiki/descriptive-statistics-and-graphics\#frequency-tables">Frequency Tables [STHDA]. </a>
* <a href="https://www.r-bloggers.com/r-function-of-the-day-table/">R Function of the Day: table [R-bloggers]. <a>
* <a href="http://www.sthda.com/english/wiki/bar-plots-r-base-graphs">Bar Plots - R Base Graphs [STHDA]. </a>
* <a href="https://www.statmethods.net/graphs/bar.html">Bar Plots [Quick-R]. </a>

::: 


## GUIDED QUESTIONS {#g13}

##### Question 7
For the marital status variable (`marital0`), what does the category of `1` represent? How many `1`'s are there in the data set?  

##### Question 8
Using the bar plot, write down the marital status groups in descending order (from largest to smallest). 

::: {.graybox}

See \@ref(answers-1) for answers

:::

### Summarize single continuous variable

Now let's learn how to summarize a single continuous variable. The LHS variable that will be the focus in this section is the variable for age (in years) at baseline, `age`.

Recall that common descriptive methods for summarizing continuous variables include measures of center, measures of spread, percentiles, and graphical displays (e.g., boxplot, histogram).

#### Summary measures

There are two ways to calculate summary measures for a continuous variable. The first option is to compute the summary statistics for a variable one-by-one, or one-line-at-a-time. The second option allows you to compute multiple summarize statistics at once.

**Option #1**

- To create a summary statistics for a continuous variable, use the following functions, specifying the object or name of the data set, followed by a `$`, and then the variable name of interest (e.g., `lhs$age`):

```{r, eval=FALSE, echo=TRUE, comment = ""}
###Measures of center functions: mean() and median()
mean(lhs$age)
median(lhs$age)

###Measures of spread functions: sd(), var() for variance, IQR(), and 
###max()-min() for range
sd(lhs$age)
var(lhs$age)
IQR(lhs$age)
max(lhs$age) - min(lhs$age)

###Other measures: min() for minimum and max() for maximum
min(lhs$age)
max(lhs$age)
```

Some of the function names give away what they calculate (e.g., `mean()` calculates the mean), but for completeness, a brief description for each function will be presented. 

`mean()` calculates the mean of the variable,  

`median()` calculates the median of the variable, 

`sd()` calculates the standard deviation of the variable,  

`var()` calculates the variance of the variable,  

`IQR()` calculates the interquartile range of the variable,  

`min()` reports the minimum value of the variable, and  

`max()` finds the maximum value of the variable.


::: {.redbox}
**R TIP**

If a variable has any missing values, then the functions above will return $NA$ for that variable. In those situations, you will need to include a <bdi style = "font-family: 'Courier New'"> na.rm=TRUE</bdi> argument in the function (e.g., <bdi style = "font-family: 'Courier New'"> mean(object\$variable,na.rm=TRUE)</bdi>). 
<br>

For example, the LHS variable for salivary cotinine at 4th annual visit (<bdi style = "font-family: 'Courier New'"> COT4 </bdi>) has missing values. To compute the mean for this variable, you would type the following code: <bdi style = "font-family: 'Courier New'"> mean(lhs\$COT4,na.rm=TRUE)</bdi>.

:::


**Option #2**

If you wanted to compute multiple summary statistics for a continuous variable in one function, you can use either the `summary()` function again, but only on the specific variable, or the `describe()` function. 

- To compute multiple summary statistics at once for a continuous variable, use the `summary()` function, specifying the object or name of the data set, followed by a `$`, and then the variable name of interest (e.g., `lhs$age`):

```{r, eval=FALSE, echo=TRUE, comment = ""}
summary(lhs$age)
```

**OR** 

The `describe()` function is located within a package in R called the `{psych}` package. Because R is an open-sourced software, there are packages being added to R all the time. Packages contain functions, data, and documentation and allows the material to be easily shared with others. You only need to install a particular package **once** on your computer using the `install.packages()` function. 

Then, to use the functions and data in a package, you must run the `library()` function, specifying the package name, with every new R session. That is, if you exit R and then come back to it, you must first run `library(<package name>)` to use the material within that package. 

- To compute multiple summary statistics at once for a continuous variable, use the `describe()` function, from the `{psych}` package, specifying the object or name of the data set, followed by a `$`, and then the variable name of interest (e.g., `lhs$age`). Two additional useful arguments to the `describe()` function include `IQR=TRUE` to calculate the IQR of the variable and `quant=c(.25,.75)` to calculate the first and third quartile (also known as Q1 or 25th percentile and Q3 or 75th percentile, respectively):

```{r, eval=FALSE,echo=TRUE, comment = ""}
#############
# Run the install.packages() function once per package
#############
install.packages("psych")

###Access the package using the library() function
library(psych)

###Calculate multiple summary statistics using the describe() function
describe(lhs$age, IQR=TRUE, quant=c(.25,.75))
```

The summary measures of most interest from the `describe()` function include the count or number of valid cases (`n`), the mean (`mean`), the standard deviation (`sd`), the median (`median`), the minimum (`min`), the maximum (`max`), the range (`range`), the standard error (`se`) (to be introduced in a later week), the IQR (`IQR`), and first and third quartile (`Q0.25` and `Q0.75`, respectively).



#### Boxplots and histograms

- To create a boxplot of a continuous variable, use the `boxplot()` function, specifying the object or name of the data set, followed by a `$` and the variable name of interest (e.g., `lhs$age`), and other arguments if you wish to customize the bar plot (e.g., `main=` for the plot title):

```{r, eval=FALSE, echo=TRUE, comment = ""}
boxplot(lhs$age, main="Boxplot of LHS Ages")
```

- To create a histogram of a continuous variable, use the `hist()` function, specifying the object or name of the data set, followed by a `$` and the variable name of interest (e.g., `lhs$age`), and other arguments to customize the bar plot (e.g., `main=` for the plot title, `xlab=` for the x-axis label, `breaks=` for modifying the binning of the histogram):

```{r, eval=FALSE, echo=TRUE, comment = ""}
hist(lhs$age, main="Histogram of LHS Ages", xlab="Ages", breaks = 10)
```

::: {.bluebox}
**SUMMARIZE SINGLE CONTINUOUS VARIABLE**

<b>Purpose:</b> To summarize a single continuous variable using summary measures and graphical representations.
<br>
<div style="border-bottom: 1px dashed black"></div>
<br>
For additional R resources on summarizing continuous variables, see: 

* <a href="http://www.sthda.com/english/wiki/descriptive-statistics-and-graphics">Descriptive Statistics and Graphics [STHDA]</a>.
* <a href="http://www.sthda.com/english/wiki/box-plots-r-base-graphs">Box Plots - R Base Graphs [STHDA].</a>
* <a href="https://www.statmethods.net/graphs/density.html">Histograms and Density Plots [Quick-R].</a>
* <a href="http://www.sthda.com/english/wiki/histogram-and-density-plots-r-base-graphs">Histogram and Density Plots - R Base Graphs [STHDA].</a>

:::

## GUIDED QUESTIONS {#g14}

##### Question 9
What is the mean age of the participants in the LHS study? What is the median age? The standard deviation of age? The minimum and maximum age?

##### Question 10 
Using the histogram, approximately how many participants were between 30 and 35 years old at the beginning of the LHS study?

::: {.graybox}

See \@ref(answers-1) for answers

:::

## CHALLENGES {#ch1}

##### Question 11
Using the `counts` object, figure out how to create a table of proportions instead of counts for the marital status variable. [Hint: Use `prop.table()` function.] 

##### Question 12
Figure out how to make the color of bars in the bar plot for marital status, `marital0`, to be blue, and how to change the y-axis title to be "Number".

##### Question 13
Figure out how to make the color of the bars in the histogram for age, `age`, to be orange, and how to modify the binning of the histogram to your liking. 

::: {.graybox}

See \@ref(answers-chal-1) for answers

:::
