# Confidence Intervals for a Proportion in R

## Goal of Software Lesson 3
::: {.graybox}
The goal of this software lesson is to learn how to

* find $z^{*}$ values from a Standard Normal distribution for a confidence interval for a proportion, and
* compute confidence intervals for a proportion.

:::

Let's use the Lung Health Study (LHS) data set again. 

Recall that the data set and a description of data can be found on Canvas, titled *lhs.csv* and *lhsdoc.docx*, respectively.

```{r, eval=FALSE, echo=TRUE, comment = ""}
lhs<-read.csv(file = file.choose(), header = TRUE)
```

## Confidence interval for a proportion

Confidence intervals (CIs) are a way to make a inference, via interval estimates, about population parameters by using the sample statistics computed from our data. 

One can compute CIs for any summary measure (e.g., mean, proportion, difference in means, odds ratio, relative risk, slope). In this section, you will learn how to compute a CI for a proportion. There are two ways to do this: manually, by using the CI formula (via the Wald method), or using a built-in function in R (`prop.test()`). 

The LHS variable that will be the focus in this section is the variable for those who smoke a pipe at 1 year after starting treatment, `f060pipe`. The question that you will be trying to answer is: 

<center>
<i>What is an estimate of the population proportion for those who smoke a pipe at 1 year after starting treatment?</i>
</center>
<br>

**Option #1: Calculating manually**

Recall that the formula for computing CIs for a proportion using the Wald method is
<center>
$\hat{p}+/-z^{*}\sqrt{\frac{\hat{p}*(1-\hat{p})}{n}}$.
</center>

When calculating this CI manually in R and from a data set, you need to save each of the following values in objects:

- the number of those with the event in the variable (e.g., `k`),  

- the total number of non-missing values in the variable (e.g., `n`),

- the sample proportion (e.g., `p.hat`), 

- the standard error of the proportion (e.g., `SE`), and

- the degree of confidence value from the appropriate theoretical distribution (e.g., `z.star`).
<br>

- Step 1: To count the number of participants who had the event of interest, you first need to name the object that will contain the number who had the event (e.g., `k`), use the assignment operator (`<-`), and then use the `length()` function in combination to the `which()` function, specifying the variable and the value of the variable that should be counted:

```{r, eval=FALSE, echo=TRUE, comment = ""}
k<-length(which(lhs$f060pipe== 1))

#viewing the object you just created
k 
```

The `length()` function counts the number of values in the object, and the `which()` function finds the observations in the variable that match the criteria (e.g., which observations in the `f060pipe` variable have a value equal to 1, `lhs$f060pipe==1`).

- Step 2: To count the total number of non-missing values in the variable, you first need to name the object that will contain the total sample size (e.g., `n`), use the assignment operator (`<-`), and then use the `length()` function in combination to the `na.omit()` function, specifying the value of the variable that should be counted:

```{r, eval=FALSE, echo=TRUE, comment = ""}
n<-length(na.omit(lhs$f060pipe))

n
```

The `na.omit()` function removes the rows in the variable that are missing, so you would only be left with a variable with non-missing rows.

- Step 3: To calculate the sample proportion, you first need to name the object that will contain the sample proportion (e.g., `p.hat`), use the assignment operator (`<-`), and then use the objects from Steps 1 and 2, dividing `k` by `n`:

```{r, eval=FALSE, echo=TRUE, comment = ""}
p.hat<-k/n

p.hat
```

- Step 4: To calculate the standard error, you first need to name the object that will contain the standard error (e.g., `SE`), use the assignment operator (`<-`), and then use the SE formula and the objects from Steps 2 and 3:

```{r, eval=FALSE, echo=TRUE, comment = ""}
####Calculating the standard error for p-hat (saved in SE)
SE<-sqrt(p.hat*(1-p.hat)/n)

SE
```

The `sqrt()` function takes the square root of a value.

- Step 5: To find the $z^{*}$ value from a Standard Normal distribution for a $95\%$ CI, you first need to name the object that will contain the $z^{*}$ value (e.g., `z.star`), use the assignment operator (`<-`), and then use the `qnorm()` function, specifying the value of 0.975:

```{r, eval=FALSE, echo=TRUE, comment = ""}
z.star<-qnorm(0.975)
```

The percentile value of $97.5\%$ was used in the above example because you want the middle $95\%$ of the distribution, so that leaves $5\%$ in both tails, or $2.5\%$ in each of the tails. Because the `qnorm()` function gives us the value where $X\%$ lies below that value, you are looking for the $z^{*}$ value such that $97.5\%$ (i.e., $95\% + 2.5\%$) falls below it.  

::: {.redbox}
**R TIP**

If you want to find $z^{*}$ values from a Standard Normal distribution for a $99\%$ or $90\%$ CI, use a similar code as above, specifying the appropriate values in the function:
<br>

<bdi style = "font-family: 'Courier New'"> qnorm(0.995) \#$z^{*}$ for $99\%$ CI</bdi>
<br>

<bdi style = "font-family: 'Courier New'"> qnorm(0.95) \#$z^{*}$ for $90\%$ CI</bdi>

:::
<br>

- Step 6: To calculate the $95\%$ CI for a proportion, use the CI mathematical formula and the objects from Steps 3, 4, and 5:

```{r, eval=FALSE, echo=TRUE, comment = ""}
p.hat + c(-1,1)*z.star*SE
```

Because the `c(-1, 1)` was added to the equation, the output will give you two numbers: one for the lower value of the CI and one for the upper value of the CI. 

**Option #2: Built-in function**

An alternative way of computing a CI for a proportion is to use the built-in function in R: `prop.test()`.

- To compute a $95\%$ CI for a proportion using R's built-in function, use the `prop.test()` function, specifying the number of those with the event in the variable (e.g., `k`), the total number of non-missing values in the variable (e.g., `n`), and the confidence level (e.g., `conf.level=0.95`):

```{r, eval=FALSE, echo=TRUE, comment = ""}
prop.test(k, n, conf.level=0.95)
```

The `prop.test()` function supplies both a hypothesis test for a proportion as well as the CI for a proportion. 

If you did not have data and only had summary information for calculating the CI for a proportion, you can just add those numbers to the function rather than saving them to an object. For example, you could have just been given the summary information from the `f060pipe` variable: 59 LHS participants who smoked a pipe at year 1 after starting treatment and 5597 participants with non-missing values in that variable. 

Try it out!
```{r, eval=FALSE, echo=TRUE, comment = ""}
prop.test(59, 5597, conf.level=0.95)
```

::: {.bluebox}
**CONFIDENCE INTERVAL FOR A PROPORTION**

<b>Purpose:</b> To estimate the population proportion using sample data via an interval estimate.
<br>
<div style="border-bottom: 1px dashed black"></div>
<br>

For additional R resources on calculating a confidence interval for a proportion, see: 

* <a href="https://www.theanalysisfactor.com/r-tutorial-count/">Counting Elements in a Data Set [The Analysis Factor].</a>
* <a href="http://www.r-tutor.com/elementary-statistics/interval-estimation/interval-estimate-population-proportion">Interval Estimate of Population Proportion [R Tutorial].</a>
* <a href="http://sphweb.bumc.bu.edu/otlt/MPH-Modules/BS/R/R-Manual/R-Manual11.html#headingtaglink_4">Confidence Interval for a Proportion [BU].</a>

:::

<br>

## GUIDED QUESTIONS {#g31}

##### Question 1
What proportion of the LHS participants smoked a pipe at year 1 (after starting treatment)?

##### Question 2
What is the $z^{*}$ value for a $95\%$ confidence interval for a proportion? 

##### Question 3
What is the $95\%$ confidence interval for the population proportion who smoke at year 1 (after starting treatment)?

::: {.graybox}

See \@ref(answers-3) for answers

:::

## CHALLENGES {#ch3}

##### Question 4
Compute a $95\%$ confidence interval for the population proportion for those who the use of nicotine gum at first annual visit after starting treatment (`AV1GUM`).

::: {.graybox}

See \@ref(answers-chal-3) for answers

:::
