# Confidence Intervals for a Mean in R

## Goal of Software Lesson 4
::: {.graybox}
The goal of this software lesson is to learn how to

* find $t^{*}$ values from a t distribution for a confidence interval for a mean, and
*  compute confidence intervals for a mean.

:::
<br>
Let's use the Lung Health Study (LHS) data set again. 

Recall that the data set and a description of data can be found on Canvas, titled *lhs.csv* and *lhsdoc.docx*, respectively.

```{r, eval=FALSE, echo=TRUE, comment = ""}
lhs<-read.csv(file = file.choose(), header = TRUE)
```

## Confidence interval for a mean

In a previous software lesson, you learned how to compute a confidence interval (CI) for a proportion. You are now going to extend that knowledge to computing a confidence interval for a mean.

Similar to the previous software lesson, you will learn two ways to compute a CI for a mean: manually, by using the CI formula, or using a built-in function in R (`ci()`). 

The LHS variable that will be the focus in this section is the variable for weight change (kg) from baseline to 5th annual visit, `wgtchg05`. The question that you will be trying to answer is: 
<center>
<i>What is an estimate of the population mean weight change from baseline to final visit after starting treatment?</i>
</center>

**Option #1: Calculating manually**

Recall that the formula for computing CIs for a mean is
<center>
$\bar{x}+/-t^{*}(\frac{s}{\sqrt{n}})$. 
</center>

When calculating this CI manually and from a data set, you need to save each of the following values in objects:

- the sample mean of the variable (e.g., `xbar`),  

- the sample standard deviation of the variable (e.g., `sample.sd`),

- the total number of non-missing values in the variable (e.g., `n`),

- the standard error of the mean (e.g., `SE`), and

- the degree of confidence value from the appropriate theoretical distribution (e.g., `t.star`).

<br>

- Step 1: To calculate the sample mean of the variable, you first need to name the object that will contain mean value (e.g., `xbar`), use the assignment operator (`<-`), and then use the `mean()` function, specifying the variable of interest:

```{r, eval=FALSE, echo=TRUE, comment = ""}
xbar<-mean(lhs$wgtchg05,na.rm=TRUE)

#viewing the object you just created
xbar
```

- Step 2: To calculate the sample standard deviation of the variable, you will carry out a similar process as in Step 1 (e.g., name the object, assignment operator, use the `sd()` function on the variable of interest):

```{r, eval=FALSE, echo=TRUE, comment = ""}
sample.sd<-sd(lhs$wgtchg05,na.rm=TRUE)

sample.sd
```

- Step 3: To count the total number of non-missing values in the variable, you first need to name the object that will contain the total sample size (e.g., `n`), use the assignment operator (`<-`), and then use the `length()` function in combination to the `na.omit()` function, specifying the value of the variable that should be counted:

```{r, eval=FALSE, echo=TRUE, comment = ""}
n<-length(na.omit(lhs$wgtchg05))

n
```

- Step 4: To calculate the standard error, you first need to name the object that will contain the standard error (e.g., `SE`), use the assignment operator (`<-`), and then use the SE formula and the objects from Steps 2 and 3:

```{r, eval=FALSE, echo=TRUE, comment = ""}
####Calculating the standard error for p-hat (saved in SE)
SE<-sample.sd/sqrt(n)

SE
```

Recall that the `sqrt()` function takes the square root of a value.

- Step 5: To find the $t^{*}$ value from a *t* distribution with $n-1$ degrees of freedom for a $95\%$ CI, you first need to name the object that will contain the $t^{*}$ value (e.g., `t.star`), use the assignment operator (`<-`), and then use the `qt()` function, specifying the value of 0.975 and the appropriate degrees of freedom (e.g., $n-1$):

```{r, eval=FALSE, echo=TRUE, comment = ""}
t.star<-qt(0.975, df=n-1)

t.star
```

::: {.redbox}
**R TIP**

If you want to find $t^{*}$ values from a <i>t</i> distribution with $n-1$ degrees of freedom for a $99\%$ or $90\%$ CI, use a similar code as above, specifying the appropriate values in the function:
<br>
<bdi style = "font-family: 'Courier New'"> qt(0.995, df=value) #$t^{*}$ for $99\%$ CI</bdi>
<br>
<bdi style = "font-family: 'Courier New'"> qt(0.95, df=value) #$t^{*}$ for $90\%$ CI</bdi>

:::

<br> 

- Step 6: To calculate the $95\%$ CI for a mean, use the CI mathematical formula and the objects from Steps 1, 4, and 5:

```{r, eval=FALSE, echo=TRUE, comment = ""}
xbar + c(-1,1)*t.star*SE
```

**Option #2: Built-in function**

An alternative way of computing a CI for a mean is to use the built-in function in R: `ci()` from the `{epiDisplay}` package.

- To compute a $95\%$ CI for a mean using R's built-in function, use the `ci()` function from the `{epiDisplay}` package, specifying the variable of interest and the alpha level (e.g., `alpha=0.05`):

```{r, eval=FALSE, echo=TRUE, comment = ""}
#############
# Run the install.packages() function once per package
#############
install.packages("epiDisplay")

###Access the package using the library() function
library(epiDisplay)

###Calculate a confidence interval using the ci() function
ci(lhs$wgtchg05, alpha=0.05)
```

The output for this function provides the sample size for the variable (non-missing values, `n`), the mean of the variable (`mean`), the standard deviation of the variable (`sd`), the standard error of the statistic (`se`), and then the lower and upper confidence values of the interval (`lower95ci` and `upper95ci`, respectively).

The `alpha=` argument specifies the confidence level, where confidence level = $1-\alpha$. For example, for a $95\%$ confidence interval, you would specify $\alpha$ as $1-0.95=0.05$. The default setting is `alpha=0.05` so you do not have to include this argument if you are computing a $95\%$ CI. 

::: {.bluebox}
**CONFIDENCE INTERVAL FOR A MEAN**

<b>Purpose:</b> To estimate the population mean using sample data via an interval estimate.
<br>
<div style="border-bottom: 1px dashed black"></div>
<br>

For additional R resources on calculating a confidence interval for a mean, see: 

* <a href="https://www.cyclismo.org/tutorial/R/confidence.html">Calculating Confidence Intervals [Cyclismo].</a>
* <a href="https://bookdown.org/danieljcarter/r4steph/constructing-confidence-intervals.html">Constructing Confidence Intervals [R for STEPH].</a>
* <a href="http://www.r-tutor.com/elementary-statistics/interval-estimation/interval-estimate-population-mean-unknown-variance">Interval Estimate of Population Mean with Unknown Variance [R Tutorial].</a>

:::
<br>

## GUIDED QUESTIONS {#g41}

##### Question 1
What is the average weight change from baseline to 5th annual visit for the LHS participants? What is the value of the SE? 

##### Question 2
What is the degrees of freedom that was used to find the $t^{*}$ value in `qt(0.975, df=n-1)`?

##### Question 3
What is the $95\%$ confidence interval for the population mean weight change from baseline to 5th annual visit?

::: {.graybox}

See \@ref(answers-4) for answers

:::

## CHALLENGES {#ch4}

##### Question 4
Compute a $95\%$ confidence interval for the population mean number of cigarettes smoked at the 5th (and last) annual visit (`AV5CIGS`).

::: {.graybox}

See \@ref(answers-chal-4) for answers

:::
