# Correlation and Simple Linear Regression in R

## Goal of Software Lesson 12

::: {.graybox}

The goal of this software lesson is to learn how to

* create a scatterplot,
* obtain a correlation coefficient,
* obtain regression coefficients (i.e., intercept and slope) and $R^2$ for a simple linear regression model, and 
* subset the data.

:::

<br>

Let's use the Lung Health Study (LHS) data set again. 

Recall that the data set and a description of data can be found on Canvas, titled *lhs.csv* and *lhsdoc.docx*, respectively.

```{r, eval=FALSE, echo=TRUE, comment = ""}
lhs<-read.csv(file = file.choose(), header = TRUE)
```

## Relationship between two numerical variables

Thus far, you have examined the relationship between a numerical variable and a categorical variable (e.g., two-sample *t*-test, ANOVA) and also between two categorical variables (e.g., Chi-square test, Fisher's exact test, McNemar's test). Now you are going to investigate the relationship between two numerical variables (e.g., scatterplot, correlation, simple linear regression). 

In the LHS study, suppose the researchers wanted to examine the relationship between the number of cigarettes smoked per day at baseline (`f10cigs`) and FEV1/FVC$\%$ at baseline (`FEVFVC02`). This will be the scenario used throughout this section of the lesson. 

### Scatterplot

The first step when examining the relationship between two numerical variables is to make a graphical display called a scatterplot. 

- To make a scatterplot, use the `plot()` function, specifying the x-variable first and then the y-variable second, separated by a `,` (e.g., `plot(x, y)`), and other arguments if you wish to customize the scatterplot (e.g., `xlab=` to specify the x-axis label, `ylab=` to specify the y-axis label, `main=` to specify the plot title):

```{r, eval=FALSE, echo=TRUE, comment = ""}
plot(lhs$f10cigs, lhs$FEVFVC02, 
     xlab="Cigarettes per day", 
     ylab="FEV/FVC%", 
     main= "FEV/FVC% vs. Cigarettes at Baseline")
```

::: {.bluebox}
**SCATTERPLOT**

<b>Purpose:</b> To graphically display the relationship between two numerical variables.
<br>
<div style="border-bottom: 1px dashed black"></div>
<br>

For additional R resources on scatterplots, see: 

* <a href"https://www.statmethods.net/graphs/scatterplot.html">Scatterplots [Quick-R].</a>
* <a href="http://www.sthda.com/english/wiki/scatter-plots-r-base-graphs">Scatterplots - R Base Graphs [STHDA].</a>
* <a href="http://www.sthda.com/english/wiki/wiki.php?title=ggplot2-scatterplot-easy-scatter-plot-using-ggplot2-and-r-statistical-software">Easy scatterplot using ggplot2 [STHDA]</a> [Note. The {ggplot2} package is a very useful package for creating elegant and aesthetically pleasing visualizations].

:::
<br>


### Correlation

If the relationship between the two numerical variables looks approximately linear, you can quantify the relationship by computing a correlation coefficient, *r*. 

- To compute the correlation coefficient, use the `cor()` function, specifying the x- and y-variables, separated by a `,`, and the argument to delete cases that have a mission value in either of the two variables (`use="complete.obs"`):

```{r, eval=FALSE, echo=TRUE, comment = ""}
cor(lhs$f10cigs, lhs$FEVFVC02, use="complete.obs")
```

The `use="complete.obs"` argument is required in the `cor()` function if there are missing values in either of the two variables. Without it, R would output `NA` for the correlation, which you know isn't true. If there are not missing values in either of those two variables, then you can leave this argument out of the function.

::: {.bluebox}
**CORRELATION**

<b>Purpose:</b> To quantify the relationship (strength and direction) between two numerical variables.
<br>
<div style="border-bottom: 1px dashed black"></div>
<br>

For additional R resources on correlation, see: 

* <a href="http://www.sthda.com/english/wiki/correlation-test-between-two-variables-in-r">Correlation Test Between Two Variables in R [STHDA].</a>
* <a href="http://sphweb.bumc.bu.edu/otlt/MPH-Modules/BS/R/R-Manual/R-Manual18.html#headingtaglink_5">Correlation [BU SPH].</a>

:::

## GUIDED QUESTIONS {#g121}

##### Question 1
What is the value of the correlation between the number of cigarettes smoked per day at baseline and FEV/FVC$\%$?

::: {.graybox}

See \@ref(answers-12) for answers

:::

### Simple linear regression

Besides correlation, another way to describe the relationship between two numerical variables is through simple linear regression (SLR) if the relationship looks approximately linear. In SLR, you obtain regression coefficients (i.e., intercept and slope) from the SLR model to predict Y from X. For example, you can create a SLR model to predict FEV1/FVC$\%$ at baseline (Y) from the number of cigarettes smoked per day at baseline (X).

- To fit a simple linear regression model, first, use the `lm()` function, specifying the y-variable and the x-variable, separated by `~` (e.g., `lm(y~x)`) and saving the information under an object name (e.g., `model1`). Then use the `summary()` function on the saved `lm` object (`model1`) to obtain the regression coefficient values: 

```{r, eval=FALSE, echo=TRUE, comment = ""}
# fitting a SLR line
model1<-lm(lhs$FEVFVC02~lhs$f10cigs)

# obtaining results in SLR object, model1
summary(model1)
```

The main information of interest is in the table under the header `Coefficients:`. 

The regression coefficient values are under the column `Estimate`, with the intercept denoted in the first row (`(Intercept)`) and the slope denoted in the second row. The information in this column is of most interest this week. 

The standard errors for each of the regression coefficients are under the column `Std. Error`. This value is used to create confidence intervals and *t*-test statistics for the coefficients.

In the third column, `t value` represents the *t*-test statistics for testing the regression coefficients. 

And lastly, the *p*-values for testing the regression coefficients are presented in the column `Pr(>|t|)`. These last three pieces of information will be discussed next week.

Additional model statistics that may be of interest include $R^2$, which is denoted next to `Multiple R-squared`, and  $\hat{\sigma}$, which is denoted next to `Residual standard error`. $R^2$ is a measure of how well the model fits the data and the residual standard error, $\hat{\sigma}$, is the standard deviation of the residual values.

### Draw line of best fit

The saved `lm` object can also be used to draw a line of best fit in a scatterplot.

- To add the line of best fit to a scatterplot, first, use the `plot()` function, specifying the x- and y-variables and other arguments if you wish to customize the scatterplot, and then use the `abline()` function on the saved `lm` object:

```{r, eval=FALSE, echo=TRUE, comment = ""}
plot(lhs$f10cigs, lhs$FEVFVC02, 
     xlab="Cigarettes per day", 
     ylab="FEV/FVC%", 
     main= "FEV/FVC% vs. Cigarettes at Baseline")
abline(model1)
```

::: {.bluebox}
**SIMPLE LINEAR REGRESSION**

<b>Purpose:</b> To estimate regression coefficients to examine (and predict) the relationship between two numerical variables: a response variable, Y, and an explanatory variable, X.
<br>
<div style="border-bottom: 1px dashed black"></div>
<br>

For additional R resources on simple linear regression, see: 

* <a href="https://www.r-bloggers.com/simple-linear-regression-2/">Simple Linear Regression [R Bloggers].</a>
* <a href="https://www.ocf.berkeley.edu/~janastas/simple-linear-regression-r.html">Doing and Interpreting Simple Linear Regression in R [Berkeley].</a>

:::
<br>

## GUIDED QUESTIONS {#g122}

##### Question 2
What is the value of the intercept in the model of `FEVFVC02` vs. `f10cigs`? What is the value of the slope? 

##### Question 3
What is the $R^2$ value for the model?

::: {.graybox}

See \@ref(answers-12) for answers

:::


## Subsetting a data set

Sometimes when you are working with a large data set, you are only interested in a small portion of it. For example, suppose the LHS researchers were only interested in the females that were in the study.

There are two ways to subset the data in R. The first option is the most basic way by using square brackets after the data name, `data[rows, columns]`. The second option (and easiest to read) is to use the `subset()` function.

### Option #1

- To subset the data set, first, create a **new** object/data set in R (e.g., `lhs.f1`), followed by the assignment operator (`<-`), then assign the new data set to the old one (e.g., `lhs`), followed by which rows you are interested in keeping (e.g., `lhs[lhs$AGENDER=="F", ]`): 

```{r, eval=FALSE, echo=TRUE, comment = ""}
lhs.f1<-lhs[lhs$AGENDER=="F", ]
```

Here's how you read the line of code:

"I want to create a new object/data set, called lhs.f1, (denoted by `lhs.f1<-`) ...
and this object is the lhs data set (`lhs[ , ]`), **but** only give me the rows for when AGENDER=="F" (denoted by the `lhs$AGENDER=="F"` before the comma in the square brackets), and give me all of the columns (denoted by the empty space after the comma in the square brackets)."

### Option #2

- To subset the data set, first, create a **new** object/data set in R (e.g., `lhs.f2`), followed by the assignment operator (`<-`), then use the `subset()` function, specifying the name of the data set first and then how you want to subset the data (e.g., `subset(data, data$variable == value`): 

```{r, eval=FALSE, echo=TRUE, comment = ""}
lhs.f2<-subset(lhs, lhs$AGENDER=="F")
```

Note: For either subsetting option, if you want to subset the data on a range of values for a numerical variable (e.g., age > 50) rather than on a single value (as presented above), then you can just supply that particular criteria in the code.

::: {.bluebox}
**SUBSETTING A DATA SET**

<b>Purpose:</b> To filter the data set based on specific criteria.
<br>
<div style="border-bottom: 1px dashed black"></div>
<br>

For additional R resources on subsetting a data set, see: 

* <a href="https://www.r-bloggers.com/5-ways-to-subset-a-data-frame-in-r/">5 Ways to Subset a Data Frame in R [R Bloggers].</a>
* <a href="https://stats.idre.ucla.edu/r/modules/subsetting-data/">Subsetting Data [UCLA].</a>

:::

<br>

## CHALLENGE QUESTIONS {#ch12}

##### Question 4
Investigate the relationship between the number of cigarettes smoked per day at baseline (`f10cigs`; Y) and the number of alcoholic drinks per week at baseline (`DRINKSBL`; X). Create a scatterplot, calculate a correlation coefficient, and obtain regression coefficients and $R^2$ from a simple linear regression model.

##### Question 5
Investigate the relationship between the FEV1/FVC$\%$ at baseline (`FEVFVC02`; Y) and the number of cigarettes smoked per day at baseline (`f10cigs`; X) for the data set that contains only females (`lhs.f1` or `lhs.f2`). Create a scatterplot, calculate a correlation coefficient, and obtain regression coefficients and $R^2$ from a simple linear regression model.

##### Question 6
Install and load the `{ggplot2}` package to create a scatterplot of the number of cigarettes smoked per day at baseline (`f10cigs`) and FEV1/FVC$\%$ at baseline (`FEVFVC02`). See the following website for tips on how to code using `ggplot2`: <a href="http://www.sthda.com/english/wiki/ggplot2-scatter-plots-quick-start-guide-r-software-and-data-visualization">ggplot2 scatterplots: Quick start guide - R software and data visualization</a>. Try modifying the size of the points, modifying the shape and color of the points, and adding the line of best fit. 

::: {.graybox}

See \@ref(answers-chal-12) for answers

:::


