# R Answer Key {-}

## Software Lesson 1 Guided Questions {#answers-1}

::: {.yellowbox}

Return to Chapter 1:

- Guided Questions 1-3 (\@ref(g11))
- Guided Questions 4-6 (\@ref(g12))
- Guided Questions 7-8 (\@ref(g13))
- Guided Questions 9-10 (\@ref(g14))
- Challenge Questions (\@ref(ch1))

:::
<br>

1. 44 years old
2. Race is given as an integer <br>
Aclinic is given as a character <br>
BMI is given as a number <br>
3. The data types match for ACLINIC and BMI, as the clinic is a factor variable and bmi should be a number.  However, race should be a character variable, not an integer. Since the different races are categorized by the numbers 1-5, R reads them as integers, not classifications.
4. Min = 2 <br> Q1 = 20 <br> Med = 30 <br> Q3 = 40 <br> Max = 120 <br> Mean = 31.27
5. Weight change (wgtchg02) and Salivary cotinine (COT4) have missing values.
6. No, because the races are categories, not numbers.
7. The category 1 represents those who have never been married, and there are 288 such participants.
8. Married, Divorced, Never Been Married, Widowed, Single
9. Mean: 48.47 years <br> Median: 49 years <br> Std. Dev: 6.83 years <br> Min: 34 years <br> Max: 67 years
10. Around 150 participants

#### Challenge Questions 1 {#answers-chal-1}

11. `prop.table(counts)`
12. 

```{r, echo = TRUE, eval= FALSE}
barplot(counts, main="Barplot of LHS Marital Status", ylab = "Number",
        names.arg=c("Never","M","W","S","D"), col = "blue")
```

13. One possible answer for this is:

```{r, eval = FALSE, echo = TRUE}
hist(lhs$age, main="Histogram of LHS Ages",
     xlab="Ages", breaks = 10, col = "orange")
```






## Software Lesson 2 Guided Questions {#answers-2}

::: {.yellowbox}

Return to Chapter 2:

- Guided Questions 1-2 (\@ref(g21))
- Guided Questions 3-5 (\@ref(g22))
- Challenge Questions (\@ref(ch2))

:::
<br>

1. 0.045
2. 0.744
3. 0.022
4. 0.030
5. Because we want to subtract those less than 1, not including 1

#### Challenge Questions 2 {#answers-chal-2}

6. Two ways to find this are:
```{r, echo = TRUE, eval = FALSE}
1 - pnorm(45, mean = 50, sd = 10)
1 - pnorm(-0.5)
```

7. 0.745
8. 0.9999693
9. 0.9999969






## Software Lesson 3 Guided Questions {#answers-3}

::: {.yellowbox}

Return to Chapter 3:

- Guided Questions 1-3 (\@ref(g31))
- Challenge Questions (\@ref(ch3))

:::
<br>

1. 0.01054
2. 1.959964
3. Manual calculation result (Wald method): (0.00787, 0.01322)

`prop.test()` function result (Wilson method): (0.00810, 0.01367)

#### Challenge Questions 3 {#answers-chal-3}

4. Manual calculation result (Wald method): (0.15806, 0.17766)

`prop.test()` function result (Wilson method): (0.15820, 0.17798)






## Software Lesson 4 Guided Questions {#answers-4}

::: {.yellowbox}

Return to Chapter 4:

- Guided Questions 1-3 (\@ref(g41))
- Challenge Questions (\@ref(ch4))

:::
<br>

1. The average weight change is 3.71064 kg, with a standard error of 0.09035.
2. 5468 degrees of freedom
3. (3.53, 3.89)

#### Challenge Questions 4 {#answers-chal-4}

4. (13.54004, 14.31612)







## Software Lesson 5 Guided Questions {#answers-5}

::: {.yellowbox}

Return to Chapter 5:

- Guided Questions 1-3 (\@ref(g51))
- Guided Questions 4-8 (\@ref(g52))
- Guided Questions 9-11 (\@ref(g53))
- Challenge Questions (\@ref(ch5))

:::
<br>

1. 3.7106 kg
2. The test statistic is $t = 41.068$ with 5468 degrees of freedom.
3. $2.2 \times 10^{-16},$ which is the smallest value R will show.
4. The mean difference is 212.97. Since the result was negative, the salivary cotinine level was higher at baseline on average.
5. No they do not.  They are the default arguments in R.
6. (202.19, 223.76)
7. The test statistic is $t = -38.73$, with 2545 degrees of freedom and a *p*-value of $p < 2.2 \times 10^{-16}$. 
8. We have already taken the differences between the two visits, so they have already been paired, and now we only have one measurement.
9. There are quotes placed around the labels because they are characters.
10. You would need six breaks for five categories.
11. "pipe1"

#### Challenge Questions 5 {#answers-chal-5}

12. Using a one-sided t-test, we have  a test statistic of $t= 32.854$, with a *p*-value of  $p = 2.2\times 10^{-16}$.
13. One way to code this is 
```{r, echo = T, eval = F}
lhs$wgt.fxn<-cut(lhs$wgtchg01,
       breaks=c(-30, -.0001, .0001, 35),
      labels=c("lost wt","no change", "gained wt"))
```






## Software Lesson 6 Guided Questions {#answers-6}

::: {.yellowbox}

Return to Chapter 6:

- Guided Questions 1-3 (\@ref(g61))
- Guided Question 4 (\@ref(g62))
- Guided Questions 5-6 (\@ref(g63))
- Guided Question 7 (\@ref(g64))
- Challenge Questions (\@ref(ch6))

:::
<br>

1. The difference in mean CO values at baseline is 3.64, where males had the higher baseline mean.
2. (-5.24, -2.04)
3. The test statistic is $t = -4.4595$ with  847.97 degrees of freedom and a *p*-value of $p = 9.32\times 10^{-6}$.
4. The test statistic is $W = 717030$ with a *p*-value of $p= 0.08908$.
5. The median age is 49 in each group.  The means are 48.41 in the SI-A group, 48.56 in the SI-P group, and 48.44 in the UC group.  The SDs are 6.84 in the SI-A group, 6.83 in the SI-P group, and 6.8 in the UC group.
6. Based on the statistics and boxplots, the groups have almost the same mean and shape, but the UC group has a smaller maximum. The histograms look almost the same for all three groups.
7. The value in the first row is 2.3026.

#### Challenge Questions 6 {#answers-chal-6}

8. One way of coding this is
```{r, echo = T, eval = FALSE}
lhs$cigdiff <- lhs$AV5CIGS - lhs$f10cigs
```

9. One way of computing these statistics is
```{r, echo = T, eval = F}
describeBy(lhs$cigdiff, lhs$AGENDER)
```

10. A histogram can be created using the following code:
```{r, echo = T, eval = F}
histogram(~lhs$cigdiff | lhs$AGENDER,
          xlab="Cigarette Difference", main="Histograms of Cig Difference by \n Sex")
```

A boxplot can be created using the following code:
```{r, echo = T, eval = F}
boxplot(lhs$cigdiff~lhs$AGENDER, main="Boxplot of Cig Differences by \n Sex", xlab = "Sex", ylab = "Cigarette Differences")
```

11. The appropriate method here is a two-sample t-test, which can be found using the following:
```{r, echo = T, eval = F}
t.test(lhs$cigdiff~lhs$AGENDER, conf.level=0.95,
       mu=0, alternative="two.sided")
```






## Software Lesson 7 Guided Questions {#answers-7}

::: {.yellowbox}

Return to Chapter 7:

- Guided Question 1 (\@ref(g71))
- Guided Questions 2-3 (\@ref(g72))
- Challenge Questions (\@ref(ch7))

:::
<br>

1. The sample size is $n = 100$ individuals per group.
2. The power is 0.692.
3. The power is 0.364.

#### Challenge Questions 7 {#answers-chal-7}

4. Three choices of effect sizes and their corresponding sample sizes are: 
<br>1. $d = 0.1, N = 1571$ 
<br>2. $d = 0.6, N = 45$
<br>3. $d = 1.2, N = 12$
5. Three choices of power and their corresponding sample sizes are: 
<br>1. $pwr = 0.7, N = 79$ 
<br>2. $pwr = 0.85, N = 114$
<br>3. $pwr = 0.95, N = 164$
6. Three choices of significance levels and their corresponding sample sizes are: 
<br>1. $\alpha = 0.1, N = 78$ 
<br>2. $\alpha = 0.05, N = 100$
<br>3. $\alpha = 0.01, N = 148$
7. Three choices of sample sizes and their corresponding powers are: 
<br>1. $N = 10$ has a power of 0.135 
<br>2. $N = 50$ has a power of 0.508
<br>3. $N = 134$ has a power of 0.904
8. Three choices of effect sizes and their corresponding powers are: 
<br>1. $d = 0.1$ has a power of 0.073 
<br>2. $d = 0.6$ has a power of 0.755
<br>3. $d = 1.2$ has a power of 1
9. Three choices of significance level and their corresponding powers are: 
<br>1. $\alpha = 0.1$ has a power of 0.55 
<br>2. $\alpha = 0.05$ has a power of 0.424
<br>3. $\alpha = 0.01$ has a power of 0.205




## Software Lesson 8 Guided Questions {#answers-8}



## Software Lesson 9 Guided Questions {#answers-9}

::: {.yellowbox}

Return to Chapter 9:

- Guided Question 1 (\@ref(g91))
- Guided Questions 2-4 (\@ref(g92))
- Challenge Questions (\@ref(ch9))

:::
<br>

1. The *F*-statistic is $F = 32.94$ with 2 and 5466 degrees of freedom and a *p*-value of $p = 6.02 \times 10^{-15}$.
2. The difference is -1.5378 (UC - SI-A), where SI-A has the higher average weight change than UC.
3. The confidence interval is (-2.054, -1.022) with a *p*- value of 0.0000000 (Note: *p*-values can't be 0, so seeing a value like this just means it's a **really** small value.)
4. The *p*-value is $p = 4.8 \times 10^{-12}$.

#### Challenge Questions 9 {#answers-chal-9}

5. One way of coding this is
```{r, echo = T, eval = FALSE}
lhs$cigdiff <- lhs$AV5CIGS - lhs$f10cigs
```

6. One way of computing these statistics is
```{r, echo = T, eval = F}
library(psych)
describeBy(lhs$cigdiff, lhs$alphagroup)
```

7. A histogram can be created using the following code:
```{r, echo = T, eval = F}
histogram(~lhs$cigdiff | lhs$alphagroup,
          xlab="Cigarette Difference", main="Histograms of Cig Difference by \n Treatment Group")
```

A boxplot can be created using the following code:
```{r, echo = T, eval = F}
boxplot(lhs$cigdiff~lhs$alphagroup, main="Boxplot of LHS Cig Differences by \n Treatment Group", xlab = "Group", ylab = "Cigarette Differences")
```

8. We would use an ANOVA to answer this research question:
```{r, echo = T, eval = F}
diff.ANOVA<-aov(lhs$cigdiff~lhs$alphagroup)
summary(diff.ANOVA)
```
Post-hoc analysis can be done using one of the following methods:
```{r, echo = T, eval = F}
TukeyHSD(diff.ANOVA)
pairwise.t.test(lhs$cigdiff, lhs$alphagroup, p.adj = "bonferroni")
```






## Software Lesson 10 Guided Questions {#answers-10}

::: {.yellowbox}

Return to Chapter 10:

- Guided Questions 1-3 (\@ref(g101))
- Guided Questions 4-6 (\@ref(g102))
- Challenge Questions (\@ref(ch10))

:::
<br>

1. The risk difference is 0.027, with a confidence interval of (0.006, 0.048).
2. The relative risk is 1.171, with a confidence interval of (1.041, 1.318).
3. The odds ratio is 1.21, with a confidence interval of (1.049, 1.396).
4. We have 7 Asian participants in this group, which is 0.0012 of the total.
5. 0.0388 of the SI-A group were Black.
6. 0.7143 of Native Americans were in the UC group.

#### Challenge Questions 10 {#answers-chal-10}

7. The new variable can be created using the following:
```{r, eval = F, echo = T}
lhs$wgt5.c<-cut(lhs$wgtchg05,
                 breaks=c(-100, -.0001, 100),
                 labels=c("lost wt","gained wt"))
```

Now we can look at the values in a table, and then put them in the proper order using the following:
```{r, eval = F, echo = T}
table(lhs$AGENDER, lhs$wgt5.c)

mytable1<-matrix(
  c(518,1530,995,2426),
  nrow=2, 
  ncol=2, 
  byrow=TRUE, 
  dimnames=list(c("Female", "Male"),
                c("Lost Weight", "Gained Weight")))
```

One possible graphical summary is given by

```{r, eval = F, echo = T}
barplot(mytable1, main="Barplot of Weight Change by Gender",
        xlab="Weight Change Group",
        col=c("darkblue","green"))

legend(x=0.2,y=3000,legend=rownames(mytable1),
      fill=c("darkblue","green"), 
      cex=0.5)
```

Then we can find estimates for RR, RD, and OR using
```{r, eval = F, echo = T}
summary(epi2x2(mytable1))
```

8. First we need to create the table, then calculate the statistics:
```{r, eval = F, echo = T}
mytable2<-matrix(
  c(111,2110,189,2034), 
  nrow=2, 
  ncol=2, 
  byrow=TRUE, 
  dimnames=list(c("Trt", "Placebo"),
                c("Fatal HA", "No Fatal HA")))

summary(epi2x2(mytable2))
```




## Software Lesson 11 Guided Questions {#answers-11}

::: {.yellowbox}

Return to Chapter 11:

- Guided Questions 1-4 (\@ref(g111))
- Guided Question 5 (\@ref(g112))
- Challenge Questions (\@ref(ch11))

:::
<br>

1. 387 females
2. The test statistic is $\chi^2 = 6.8252$, with one degree of freedom, and a *p*-value of $p = 0.008988$.
3. 351.6661 females
4. $p = 0.00964$
5. The test statistic is $\chi^2 = 6.5455$ with one degree of freedom and a *p*-value of $p = 0.0105$.

#### Challenge Questions 11 {#answers-chal-11}

6. First, we can create our table with the following:
```{r, echo = T, eval = F}
newtable <- table(lhs$f31pipe, lhs$alphagroup)
rownames(newtable) <- c("Pipe Smoker", "Not Pipe Smoker")
```

Now, we create a plot:
```{r, echo = T, eval = F}
barplot(newtable, beside=TRUE,
        main="Barplot of Pipe Smoking at Baseline by Intervention Group",
        xlab="Intervention Group",
        col=c("darkblue","red"))
legend(x=6.25,y=1500,legend=rownames(newtable),
       fill=c("darkblue","red"),
       cex=0.5)
```

One possibility here is to use the chi-squared test:
```{r, echo = T, eval = F}
newchi.test<-chisq.test(newtable, correct=FALSE)
newchi.test
```

7. First, we can view our variables and then create  with the following:
```{r, echo = T, eval = F}
table(lhs$A5CIGAR, lhs$BLCIGAR)

mytable<-matrix(
  c(43,69, 214,5228), 
  nrow=2, 
  ncol=2,
  byrow=TRUE, 
  dimnames=list(c("Smoked Cigar 5th", "No Cigar 5th"),
                c("Cigar at BL", "No Cigar at BL")))
```

Then we can run a McNemar's test:

```{r, echo = T, eval = F}
mcnemar.test(mytable)
```






## Software Lesson 12 Guided Questions {#answers-12}

::: {.yellowbox}

Return to Chapter 12:

- Guided Question 1 (\@ref(g121))
- Guided Questions 2-3 (\@ref(g122))
- Challenge Questions (\@ref(ch12))

:::
<br>

1. $r = -0.0378$
2. The intercept is 65.5043, and the slope is -0.0180.
3. The $R^2$ value is 0.001432.

#### Challenge Questions 12 {#answers-chal-12}

4. The following can be used to find the correlation, create the model, and create a scatterplot for the model:

```{r, echo=T, eval = F}
cor(lhs$DRINKSBL,lhs$f10cigs, use = "complete.obs")

model3 <- lm(lhs$f10cigs~lhs$DRINKSBL)
summary(model3)

plot(lhs$DRINKSBL, lhs$f10cigs, 
     xlab="Drinks at Baseline", 
     ylab="Cigarettes per day", 
     main= "Cigarettes at Baseline vs Drinks at Baseline")
abline(model3)
```

5. The following can be used to find the correlation, create the model, and create a scatterplot for the model:

```{r, echo=T, eval = F}
cor(lhs.f1$FEVFVC02,lhs.f1$f10cigs, use = "complete.obs")

model2 <- lm(lhs.f1$FEVFVC02~lhs.f1$f10cigs)
summary(model2)

plot(lhs.f1$f10cigs, lhs.f1$FEVFVC02, 
     xlab="Cigarettes per day", 
     ylab="FEV/FVC%", 
     main= "FEV/FVC% vs. Cigarettes at Baseline for Females")
abline(model2)
```

6. The following will create a scatter plot using purple triangle shapes, and fits a regression line:

```{r, echo = T, eval = F}
install.packages("ggplot2")
library(ggplot2)
ggplot(lhs, aes(x=f10cigs, y=FEVFVC02)) + 
    geom_point(size = 4, shape = 6, color = "purple") + geom_smooth(method=lm)
```







## Software Lesson 13 Guided Questions {#answers-13}

::: {.yellowbox}

Return to Chapter 13:

- Guided Questions 1-2 (\@ref(g131))
- Challenge Questions (\@ref(ch13))

:::
<br>

1. The test statistic is $t = -2.905$, with a *p*-value of $p = 0.00369$.
2. (-0.0301, -0.0058)

#### Challenge Questions 13 {#answers-chal-13}

3. The following will give estimates and confidence intervals for the model coefficients, as well as diagnostic plots: 

```{r, echo = T, eval = F}
model2 <- lm(lhs$f10cigs~lhs$DRINKSBL)
summary(model2)
confint(model2, level=0.95)
autoplot(model2)
```





## Software Lesson 14 Guided Questions {#answers-14}

::: {.yellowbox}

Return to Chapter 14:

- Guided Questions 1-3 (\@ref(g141))
- Guided Questions 4-6 (\@ref(g142))
- Challenge Questions (\@ref(ch14))

:::
<br>

1. Red and black.
2. A filled-in circle.
3. Males have the highest number of cigarettes, as more of the blue dots are on the right side of the graph.  They also have the lowest FEV1/FVC$\%$ at baseline.
4. The intercept is 65.83078, the coefficient for f10cigs is -0.01388, the coefficient for AGENDERM is -0.84248, and the coefficient for DRINKSBL is 0.01740.
5. The test statistic is $t = -2.225$, with a *p*-value of $p = 0.0261$.
6. (-1.1699, -.5151)

#### Challenge Questions 14 {#answers-chal-14}

7. One way this can be done is with the following:
```{r, echo = T, eval = F}
lhs$trtgroup <- ifelse(lhs$alphagroup == "UC", "control", "intervention")
```

8. This can be done using the following:

```{r, eval = F, echo = T}
plot(x=lhs$AGESTART, y=lhs$AV2CIGS,
     col=c("darkorange", "mediumturquoise")[factor(lhs$trtgroup)],
     pch=16,
     ylab="Number of cigarettes at 2nd Visit (per day)",
     xlab="Age started smoking")

legend("bottomright", legend=levels(factor(lhs$trtgroup)), fill=c("darkorange", "mediumturquoise"), title = "Treatment")
```

9. This can be done using the following:
```{r, echo = T, eval = F}
model3<-lm(AV2CIGS ~  factor(trtgroup) + AGESTART, data=lhs)
summary(model3)
confint(model3, level=0.95)
```
